MyTitle

### 2024-11-04 paper/main.tex
updated

To use Gunther et al.'s formalization, we adopted the c.t.m.\  approach for our proof, as in their independence proof of CH.
Specifically, we assumed the existence of a c.t.m.\  of ZF and constructed a model of ZF + $\neg$AC by forcing.
This model is known as the basic Cohen model, which is a symmetric extension of assumed c.t.m.\ 
Our proof largely follows karagila's lecture note \cite{karagila}, with some parts also referencing Jech's books \cite{jech_set_theory,jech_AC}
as these resources align well with this approach.

@@

\begin{dfn}
  The axioms of ZF are the following statements:
  \begin{itemize}
    \item Extensionality: $\forall x \forall y (\forall z (z \in x \leftrightarrow z \in y) \rightarrow x = y)$
    \item Pairing: $\forall x \forall y \exists z \forall w (w \in z \leftrightarrow w = x \lor w = y)$
    \item Union: $\forall x \exists y \forall z (z \in y \leftrightarrow \exists w (z \in w \land w \in x))$
    \item Power set: $\forall x \exists y \forall z (z \in y \leftrightarrow z \subseteq x)$
    \item Infinity: $\exists x (\emptyset \in x \land \forall y (y \in x \rightarrow y \cup \{y\} \in x))$
    \item Regularity: $\forall x(x \neq \emptyset \rightarrow \exists y (y \in x \land y \cap x = \emptyset))$
    \item Infinity: $\forall x (x \neq \emptyset \rightarrow \exists y (y \in x \land \forall z (z \in x \rightarrow z \notin y)))$
    \item Separation: $\forall p \forall x \exists y \forall z (z \in y \leftrightarrow z \in x \land \phi(z, p))$
    \item Replacement: $\forall p (\forall x \forall y \forall z (\phi(x, y, p) \land \phi(x, z, p) \rightarrow y = z) \rightarrow \\ \forall X \exists Y \forall y (y \in Y \leftrightarrow \exists x (x \in X \land \phi(x, y, p))))$
  \end{itemize}
\end{dfn}
Where separation and replacement are axiom schemas, representing infinitely many axioms for each formula $\phi$ with an appropriate arity.
\begin{dfn} 
  \emph{The axiom of choice (AC)} is the following statement: \\
  $\forall x \exists f (\mathrm{"} f \mathrm{ is a function on } x \mathrm{"} \land \forall y (y \in x \rightarrow f(y) \in y))$
\end{dfn}
Where the phrase "f is a function on x" is also considered an abbreviation in the language of set theory.
Theory ZF + AC is denoted by ZFC.
Additionally, we introduce the well-ordering theorem, as we treat AC in this form.

@@

Let $M$ be a c.t.m.\  of ZF, $\mathbb{P}$ be the set of all finite partial functions from $\omega \times \omega$ to $\{0, 1\}$, and $\leq_{\mathbb{P}}$ be $\supseteq$.
Note that the maximum element $1_{\mathbb{P}}$ is the empty set. 
Let $\pi$ be a bijection on $\omega$. $\pi$ induces an automorphism on $\mathbb{P}$ defined as follows:
$$\mathrm{dom}(\pi p) = \{ (\pi n, m) \mid (n, m) \in \mathrm{dom}(p) \}$$
$$(\pi p)(\pi n, m) = p(n, m)$$
This automorphism further induces an automorphism on $\mathbb{P}$-names. 
Let $\mathcal{G}$ be the group of all such automorphisms.
For every finite $e \subseteq \omega$, let 
$$\mathrm{fix}(e) = \{ \pi \in \mathcal{G} \mid \forall n \in e (\pi n = n) \}$$
Let $\mathcal{F}$ be the set of all subgroups $H$ of $\mathcal{G}$ such that there exists a finite $e \subseteq \omega$ with $\mathrm{fix}(e) \subseteq H$.
Note that $\mathcal{F}$ is a normal filter on $\mathcal{G}$. Let $\mathcal{N} = \mathrm{HS}^{G}$.
Since $\mathcal{N}$ is a symmetric extension of $M$, it is a c.t.m.\  of ZF.

@@

\begin{comment}
\subsection{Relativized Forcing Relation $\Vdash_{\mathrm{HS}}$}
\subsection{The Symmetric Lemma}
\subsection{Separation}
\subsection{Replacement}
\subsection{Other Axioms}
\end{comment}

\section{Defining the Basic Cohen Model}
\begin{comment}
\subsection{The Notion of Forcing}
\subsection{The Group of Automorphisms}
\subsection{The Normal Filter}
\end{comment}

\section{Proving the Basic Cohen Model satisfies $\neg$AC}

>>>>

To use Gunther et al.'s formalization, we adopted the c.t.m.\  approach for our proof, as in their independence proof of CH.
Specifically, we assumed the existence of a c.t.m.\  of ZF and constructed a model of ZF + $\neg$AC by forcing.
This model is known as the basic Cohen model, which is a symmetric extension of assumed c.t.m.\ 
Our proof is based on Karagila's lecture note \cite{karagila} and Jech's book \cite{jech_set_theory,jech_AC},
as these resources align well with this approach.

@@

\begin{dfn}
  The axioms of ZF are the following statements:
  \begin{itemize}
    \item Extensionality: $\forall x \forall y (\forall z (z \in x \leftrightarrow z \in y) \rightarrow x = y)$
    \item Pairing: $\forall x \forall y \exists z \forall w (w \in z \leftrightarrow w = x \lor w = y)$
    \item Union: $\forall x \exists y \forall z (z \in y \leftrightarrow \exists w (z \in w \land w \in x))$
    \item Power set: $\forall x \exists y \forall z (z \in y \leftrightarrow z \subseteq x)$
    \item Infinity: $\exists x (\emptyset \in x \land \forall y (y \in x \rightarrow y \cup \{y\} \in x))$
    \item Regularity: $\forall x(x \neq \emptyset \rightarrow \exists y (y \in x \land y \cap x = \emptyset))$
    \item Infinity: $\forall x (x \neq \emptyset \rightarrow \exists y (y \in x \land \forall z (z \in x \rightarrow z \notin y)))$
    \item Separation: $\forall p \forall x \exists y \forall z (z \in y \leftrightarrow z \in x \land \phi(z, p))$
    \item Replacement: $\forall p (\forall x \forall y \forall z (\phi(x, y, p) \land \phi(x, z, p) \rightarrow y = z) \rightarrow \\ \forall X \exists Y \forall y (y \in Y \leftrightarrow \exists x (x \in X \land \phi(x, y, p))))$
  \end{itemize}
\end{dfn}
Where separation and replacement are axiom schemas, representing infinitely many axioms for each formula $\phi$ with an appropriate arity.
\begin{dfn} 
  \emph{The axiom of choice (AC)} is the following statement: \\
  $\forall x \exists f (\mathrm{"} f \text{ is a function on } x \mathrm{"} \land \forall y (y \in x \rightarrow f(y) \in y))$
\end{dfn}
Where the phrase "f is a function on x" is also considered an abbreviation in the language of set theory.
Theory ZF + AC is denoted by ZFC.
Additionally, we introduce the well-ordering theorem, as we treat AC in this form.

@@

Let $M$ be a c.t.m.\  of ZF, $\mathbb{P}$ be the set of all finite partial functions from $\omega \times \omega$ to $2 = \{0, 1\}$
that are elements of $M$, and $\leq_{\mathbb{P}}$ be $\supseteq$.
Note that the maximum element $1_{\mathbb{P}}$ is the empty set. 
Let $\pi$ be a bijection on $\omega$. $\pi$ induces an automorphism on $\mathbb{P}$ defined as follows:
$$\mathrm{dom}(\pi p) = \{ (\pi n, m) \mid (n, m) \in \mathrm{dom}(p) \}$$
$$(\pi p)(\pi n, m) = p(n, m)$$
This automorphism further induces an automorphism on $\mathbb{P}$-names. 
Let $\mathcal{G}$ be the group of all such automorphisms.
For every finite $e \subseteq \omega$, let 
$$\mathrm{fix}(e) = \{ \pi \in \mathcal{G} \mid \forall n \in e (\pi n = n) \}$$
Let $\mathcal{F}$ be the set of all subgroups $H$ of $\mathcal{G}$ such that there exists a finite $e \subseteq \omega$ with $\mathrm{fix}(e) \subseteq H$.
Note that $\mathcal{F}$ is a normal filter on $\mathcal{G}$. Let $\mathcal{N} = \mathrm{HS}^{G}$.
Since $\mathcal{N}$ is a symmetric extension of $M$, it is a c.t.m.\  of ZF.

@@

\section{Constructing a Model of ZF+$\neg$AC}
In this section, we define the basic Cohen model which is a symmetric extension that does not satisfy the axiom of choice,
and prove that it is a model of ZF+$\neg$AC, as outlined in section \ref{sec:outline}.
First, we work within the locale \texttt{"M\_ctm"} which fixes a c.t.m.\  \texttt{"M"} of ZF.

\subsubsection{The Notion of Forcing}
the notion of forcing for the basic Cohen model is defined as the set of finite partial functions from $\omega \times \omega$ to $2$ that are elements of the ground model,
with the order relation given by the superset relation.
We define these as \texttt{"Fn"} and \texttt{"Fn\_leq"} in Isabelle/ZF.

\begin{isaframe}
\begin{minted}[breaklines]{isabelle}
definition Fn where "Fn ≡ { f ∈ Pow((nat × nat) × 2) ∩ M. 
                function(f) ∧ domain(f) ⊆ nat × nat ∧ finite_M(domain(f)) ∧ range(f) ⊆ 2 }"

definition Fn_leq where "Fn_leq ≡ { <f, g> ∈ Fn × Fn. g ⊆ f }" 
\end{minted}
\end{isaframe}
Where \texttt{"nat"} denotes $\omega$ and \texttt{"finite\_M(A)"} denotes that the set \texttt{"A"} is finite.
From the following lemma, this ordered set is an instance of the locale \texttt{"forcing\_data\_partial"}, 
meaning that this is indeed a notion of forcing and forms a partially ordered set with the empty set as the maximum element.
\begin{isaframe}
\begin{minted}[breaklines]{isabelle}
lemma Fn_forcing_data_partial : "forcing_data_partial(Fn, Fn_leq, 0, M, enum)" 
\end{minted}
\end{isaframe}
Where \texttt{"enum"} is the enumeration function of the set \texttt{"M"} fixed in the locale \texttt{"M\_ctm"}.

\subsubsection{The Group of Automorphisms and the Normal Filter}
Next, we define the group of automorphisms \texttt{"Fn\_perms"} on the forcing notion \texttt{"Fn"}.
\begin{isaframe}
\begin{minted}[breaklines]{isabelle}
definition nat_perms where "nat_perms ≡ bij(nat, nat) ∩ M" 
definition Fn_perm where "Fn_perm(f, p) ≡ { <<f`n, m>, l> . <<n, m>, l> ∈ p }" 
definition Fn_perm' where "Fn_perm'(f) ≡ { <p, Fn_perm(f, p)>. p ∈ Fn }" 
definition Fn_perms where "Fn_perms ≡ { Fn_perm'(f). f ∈ nat_perms }" 
\end{minted}
\end{isaframe}
Note that \texttt{"Fn\_perm'(f)"} is the automorphism induced by the bijection \texttt{"f"} on $\omega$,
and \texttt{"Fn\_perm(f, p)"} is the value of the automorphism \texttt{"Fn\_perm'(f)"} on the partial function \texttt{"p"}.
We prove that \texttt{"Fn\_perm'(f)"} is indeed an automorphism on \texttt{"Fn"} and 
\texttt{"Fn\_perms"} is a group of automorphisms on \texttt{"Fn"},
i.e., closed under composition and the inverse operation, and contains the identity function.

\begin{isaframe}
\begin{minted}[breaklines]{isabelle}
lemma Fn_perm'_is_P_auto : 
    fixes f 
    assumes "f ∈ nat_perms" 
    shows "forcing_data_partial.is_P_auto(Fn, Fn_leq, M, Fn_perm'(f))"

lemma Fn_perms_group : 
    "forcing_data_partial.is_P_auto_group(Fn, Fn_leq, M, Fn_perms)" 
\end{minted}
\end{isaframe}

Then, we define the normal filter \texttt{"Fn\_perms\_filter"} on the group of automorphisms \texttt{"Fn\_perms"},
and verify that our definition satisfies the prerequisites of the locale \texttt{"M\_symmetric\_system"}.
This completes both the definition of the basic Cohen model and
the proof that it is a model of ZF.

\begin{isaframe}
\begin{minted}[breaklines]{isabelle}
definition Fn_perms_filter where "Fn_perms_filter ≡ { 
              H ∈ forcing_data_partial.P_auto_subgroups(Fn, Fn_leq, M, Fn_perms).  
              ∃E ∈ Pow(nat) ∩ M. finite_M(E) ∧ Fix(E) ⊆ H 
    }"

lemma Fn_M_symmetric_system : 
    "M_symmetric_system(Fn, Fn_leq, 0, M, enum, Fn_perms, Fn_perms_filter)" 
\end{minted}
\end{isaframe}

Finaly, we prove that the basic Cohen model does not satisfy the axiom of choice.
To do this, we construct a set $A$ that cannot be well-ordered in it.
We work within the locale \texttt{"M\_symmetric\_system"} with parameters, as defined above, 
fixed to generate the basic Cohen model.

We formalize $a_n$ and $A$, introduced in section \ref{sec:outline}, 
as \texttt{"binmap\_row(G, n)"} and \texttt{"binmap(G)"}, respectively, for given generic filter \texttt{"G"}.
\begin{isaframe}
\begin{minted}[breaklines]{isabelle}
definition binmap_row where 
    "binmap_row(G, n) ≡ { m ∈ nat. ∃p ∈ G. p`<n, m> = 1 }"

definition binmap where "binmap(G) ≡ { binmap_row(G, n). n ∈ nat }" 
\end{minted}
\end{isaframe}
Then, we prove that $a_n$ are pairwise disjoint, which implies that $A$ is an infinite set.
\begin{isaframe}
\begin{minted}[breaklines]{isabelle}
lemma binmap_row_distinct : 
    fixes G n m 
    assumes "M_generic(G)" "n ∈ nat" "m ∈ nat" "n ≠ m" 
    shows "binmap_row(G, n) ≠ binmap_row(G, m)" 
\end{minted}
\end{isaframe}
Using this lemma, we prove that 
$p \Vdash_{\mathrm{HS}} \text{"} \dot{F} \text{ is a injection from } \dot{\omega} \text{ to } \dot{A} \text{"}$ leads a contradiction 
for any $p \in $ \texttt{"Fn"} and $\dot{F} \in $ \texttt{"HS"}.
This means that there is no injection from $\omega$ to $A$ in the basic Cohen model.
\begin{isaframe}
\begin{minted}[breaklines]{isabelle}
lemma no_injection : 
  fixes F' p0
  assumes "F' ∈ HS" "p0 ∈ Fn"  
                  "ForcesHS(p0, injection_fm(0, 1, 2), [check(nat), binmap', F'])" 
  shows False
\end{minted}
\end{isaframe}
Where \texttt{"ForcesHS(p, φ, env)"} is another notation for \texttt{"p ⊩HS φ env"},
and \texttt{"check(nat)"} and \texttt{"binmap'"} are the hereditarily symmetric names of \texttt{"nat"} and \texttt{"binmap"}, respectively.

Finally, from this lemma, we can show that assuming $A$ is well-orderable leads to a contradiction,
because $A$ is an infinite set, which implies the existence of an injection from $\omega$ to $A$.

\begin{isaframe}
\begin{minted}[breaklines]{isabelle}
lemma no_wellorder : 
    fixes r G
    assumes 
            "M_generic(G)" "wellordered(##SymExt(G), binmap(G), r)" "r ∈ SymExt(G)" 
    shows False
\end{minted}
\end{isaframe}

From the above. we complete the proof that the basic Cohen model is a model of ZF+$\neg$AC.
Our result is stated in the following theorem, with no additional assumptions from any locale.
